# Initial Concepts of Probability {#intro}

+ **Probability** is the quantification of event uncertainty. For instance, probability of getting (H)eads in a coin toss is $1/2$. Deterministic models will give the same results given the same inputs (e.g. 2 times 2 is 4), but probabilistic models might yield different outcomes.

+ An **experiment** is a process that generates data. For instance, tossing a coin is an experiment. **Outcome** is the realization of an experiment. Possible outcomes for a coin toss is Heads and Tails.

+ **Sample space** ($\mathbb{S}$) is the collection of all the possible outcomes of an experiment. Sample space of the coin toss is $\mathbb{S} = \{H,T\}$. Sample space of two coin tosses experiment is $\mathbb{S} = \{HH,HT,TH,TT\}$. Sample space can be discrete (i.e. coin tosses) as well as continuous (i.e. All real numbers between 1 and 3. $\mathbb{S} = \{x | 1 \le x \le 3, x \in \mathbb{R}\}$) _(Side note: Sample space is not always well defined.)_

+ An **event** is a subset of sample space. While outcome represents a realization, event is an information. Probability of an event $P(A)$, say getting two Heads in two coin tosses is $P(A) = 1/4$.

+ A **random variable** represents an event is dependent on a probabilistic process. On the other hand, a **deterministic variable** is either a constant or a decision variable. For instance, value of the dollar tomorrow can be considered a random variable but the amount I will invest is a decision variable (subject to no probabilistic process) and spot (current) price of the dollar is a constant.

## Set Operations

+ **Complement** of an event ($A^\prime$) with respect to the sample space represents all elements of the sample space that are not included by the event (A). For instance, complement of event $A=\{HH\}$ is $A^\prime=\{HT,TH,TT\}$

+ **Union** of two events $A$ and $B$ ($A \cup B$) is a set of events which contains all elements of the respective events. For example, say $A$ is the set that contains events which double Heads occur ($A = \{HH,HT,TH\}$) and $B$ is the set which Tails occur at least once ($B = \{TT,HT,TH\}$). The union is $A \cup B = \{HH,TH,HT,TT\}$.

+ **Intersection** of two events $A$ and $B$ ($A \cap B$) contains the common elements of the events. For example, say $A$ is the set that contains events which Heads occur at least once ($A = \{HH,HT,TH\}$) and $B$ is the set which Tails occur at least once ($B = \{TT,HT,TH\}$). The intersection is $A \cap B = \{TH,HT\}$.

+ **Mutually exclusive** or disjoint events mean that two events have empty intersection ($A \cap B = \emptyset$) and their union ($A \cup B$) contains the same amount of elements as the sum of their respective number of elements. Also $P(A \cap B) = 0$ and $P(A \cup B) = P(A) + P(B)$. For example getting double Heads ($HH$) and double Tails ($TT$) are mutually exclusive events.

## Axioms of Probability

1. Any event $A$ belonging to the sample space $A \in \mathbb{S}$ should have nonnegative probability ($P(A) \ge 0$).
2. Probability of the sample space is one ($P(\mathbb{S}) = 1$).
3. Any disjoint events ($A_i \cap A_j = \emptyset \ \forall_{i,j \in 1 \dots n}$) satisfies $P(A_1 \cup A_2 \cup \dots \cup A_n) = P(A_1) + P(A_2) + \dots  + P(A_n)$.

## Other Set and Probability Rules

+ $(A^\prime)^\prime = A$
+ $S^\prime = \emptyset$
+ $\emptyset^\prime = S$
+ $(A \cap B)^\prime = A^\prime \cup B^\prime$
+ $(A \cup B)^\prime = A^\prime \cap B^\prime$
+ $(A \cup B) \cap C = (A \cap C) \cup (B \cap C)$
+ $(A \cap B) \cup C = (A \cup C) \cap (B \cup C)$
+ $(A \cup B) \cup C = A \cup (C \cup B)$
+ $(A \cap B) \cap C =  A \cap (C \cap B)$
+ $A \cup A^\prime = \mathbb{S}$ and $A \cap A^\prime = \emptyset$ so $P(A) = 1 - P(A^\prime)$. This is especially useful for many problems. For example the probability of getting at least one Heads in a three coin tosses in a row is $1 - P(\{TTT\}) = 7/8$, the complement of no Heads in a three coin tosses in a row. Otherwise, you should calculate the following expression.

    $$P(\{HTT\}) + P(\{THT\}) + P(\{TTH\}) + P(\{HHT\}) + P(\{HTH\}) + P(\{THH\}) + P(\{HHH\}) = 7/8 $$

+ If $A \subseteq B$\ then $P(A) \le P(B)$.
+ $P(A \cup B) = P(A) + P(B) - P(A \cap B)$.
+ $P(A \cup B \cup C) =  P(A) + P(B) + P(C) - P(A \cap B) - P(B \cap C) - P(A \cap C) + P(A \cap B \cap C)$


## Counting

Counting rules will help us enumerate the sample space. It will include multiplication rule, permutation and combination.

### Multiplication Rule

If I have a series of independent events, say $1$ to $k$, and number of possible outcomes are denoted with $n_1$ to $n_k$; total number of outcomes in the sample space would be $n_1n_2\dots n_k$.

Take a series of coin tosses in a row. If I toss a coin its sample space consists of 2 elements such as $\{H,T\}$. If I toss 2 coins the sample space would be 2\*2 $\{HH,HT,TH,TT\}$. If I toss 3 coins, the sample space would be 2\*2\*2 $\{HHH,HTH,THH,TTH,HHT,HTT,THT,TTT\}$.

A poker card consists of a type and a rank. There are four types of playing cards (clubs, diamonds, hearts and spades) and 13 ranks (A - 2 to 10 - J - Q - K). Number of cards in a deck is 4\*13 = 52.

### Permutation Rule

Permutation is the arrangement of all or a subset of items.

+ Given a set of items, say $A = {a,b,c}$ in how many different ways I can order the elements? Answer is n!. In our case it is, $3! = 3.2.1 = 6$.

    $$A = \{a,b,c\},\{b,a,c\},\{b,c,a\},\{c,a,b\},\{c,b,a\},\{a,c,b\}$$

+ Suppose there are 10 (n) participants in a competition and 3 (r) medals (gold, silver and bronze). How many possible outcomes are there? Answer is $n(n-1)(n-2)\dots (n-r+1) = \dfrac{n!}{(n-r)!} = \dfrac{10!}{(10-3)!} = 720$.

+ If there are more than one same type items in a sample, then the permutation becomes $\dfrac{n!}{n_1!n_2!\dots n_k!}$ where $\sum n_i = n$.

    For example enumerate the different outcomes of four coin tosses which result in 2 heads and 2 tails. Answer is $\dfrac{4!}{2!2!} = 6$

    $$A = \{HHTT,HTTH,HTHT,THTH,THHT,TTHH\}$$

### Combination Rule

Suppose we want to select $r$ items from $n$ items and the order does not matter. So the number of different outcomes can be found using $\binom{n}{r} = \dfrac{n!}{(n-r)!r!}$.

Out of 10 students how many different groups of 2 students can we generate? Answer $\dfrac{10!}{8!2!} = 45$

## Chapter Problems

1. Suppose I toss a coin, roll a die and draw a card from the deck. How many different number of outcomes are there for this experiment?

    Solution: Multiplication rule. $n_1n_2n_3 = 2.6.52 = 624$.

    ```{r}
    n1 <- 2 #A coin toss has two potential outcomes.
    n2 <- 6 #A die roll has six potential outcomes.
    n3 <- 52 #A card draw has 52 potential outcomes.
    ```

2. In how many ways can I order the Teletubbies? (Tinky-Winky, Dipsy, Laa Laa and Po) For instance, (TW - Dipsy - Po - Laa Laa) is an ordering and (Dipsy - Po - TW - Laa Laa) is another.

    Solution: Permutation rule. $n! = 4! = 24$

    ```{r}
    n_tubbies <- 4 #Number of teletubbies
    factorial(n_tubbies) #By permtuation it is 4!
    ```

3. I want to reorder the letters of the phrase "GOODGRADES". In how many ways can I do it?.

    Solution: Remember the permutation rule with identical items. There are two "G"s, two "D"s and two "O"s. Remember the formula $\dfrac{n!}{n_1!n_2!\dots n_k!}$. So the result should be $\dfrac{10!}{2!2!2!1!1!1!1!} = 453600$.

    ```{r, tidy=TRUE}
    the_phrase <- "GOODGRADES"
    freq_table <- table(strsplit(the_phrase,split="")[[1]]) #Let's create a frequency table first
    print(freq_table) #Let's show it
    the_dividend <- factorial(nchar(the_phrase)) #Dividend part is 10 characters so 10!
    the_divisor <- prod(factorial(freq_table)) #Get multiplication of factorials for the divisor
    the_dividend/the_divisor
    ```

4. I want to make two letter words from "GRADES" such as "GA", "ED" or "DE" (it doesn't have to make sense). Find the number of permutations.

    Solution: Permutation of $r$ items from $n$ items is $\dfrac{n!}{(n-r)!}$. So the result is $\dfrac{6!}{4!} = 30$.

    ```{r}
    the_phrase<-"GRADES"
    letter_length <- 2 #We want two letter words
    #Since all letters are different no need for special permutation.
    factorial(nchar(the_phrase))/factorial(nchar(the_phrase)-letter_length)
    ```

5. Suppose I am drawing a hand of 5 cards from a playing deck of 52 cards. How many different hands there can be? (Each card is different. See the bottom of this document for details.)

    Solution: Since in a hand you do not care for the order, it is the combination $\binom{52}{5} = \dfrac{52!}{(52-5)!5!} = 2598960$.

    ```{r}
    #Combination (a.k.a binomial coefficient) function is choose
    choose(52,5)
    ```

## Extra Problems

+ Question 1
Suppose we draw three cards from a deck and roll two dice. Answer the following questions.

    a) What is the experiment?

        The experiment is "drawing three cards from a deck and rolling two dice".

    b) What is "getting two-sixes and three-kings or five-one (in any order) one queen one king one ace"? Pick one (Event / Outcome / Sample Space)

        Event.

    c) Give an example of two mutually exclusive events. (6 pts)

        Event A: Queen of Hearts / Queen of Spades / Queen of Diamonds / 6 / 5
        Event B: Ace of Clubs / King of Clubs / Queen of Clubs / 4 / 4

    d) What is the probability of getting four-three (in any order) in dice roll and three queens in card draw?

        ```{r}
        #First roll can either be 3 or 4 and second roll should be the other
        # 2/6 * 1/6
        #Getting the first Queen has probability of 4/52
        #Getting the second Queen has probability of 3/51
        #Getting the third Queen has probability of 2/50
        2/6*1/6*4/52*3/51*2/50
        ```

    e) How many different outcomes can there be? This time assume ordering is important (e.g. 6-1 and 1-6 are different outcomes).

        ```{r}
        #Six outcomes per die
        #52 outcomes for the first card draw
        #51 outcomes for the second card draw
        #50 outcomes for the second card draw
        #Multiplication rule
        #You can also use permutation rule for cards
        6*6*52*51*50
        ```

+ Question 2
    In how many ways can you arrange the letters of "HOUSEPARTY"?

    a) Any order.

        ```{r}
        the_phrase<-"HOUSEPARTY"
        #No repetitive letters
        #Permutation rule
        factorial(nchar(the_phrase))
        ```

    b) Vowels together?

        ```{r}
        #4 vowels, 6 consonants
        #Assume all vowels are a single "letter". So 8 characters.
        #But vowels permutate within the single "letter".
        #Multiplication rule
        factorial(6+1)*factorial(4)
        ```

    c) Vowels in alphabetical order?

        ```{r}
        #We start with all the permutations 10!
        #For any permutation there can be only one ordering of vowels.
        #For instance HOUSEPARTY is not valid but HAESOPURTY is valid
        #So remove invalid permutations with division
        factorial(10)/factorial(4)
        ```

    d) There should be no consecutive vowels?

        ```{r}
        #There are 6 consonants, 4 vowels.
        # Assume Xs are consonants and .s are potent vowel places.
        # .X.X.X.X.X.X.
        #Consonants can permutate in any order so 6! there
        #7 places for vowels but only 4 vowels.
        # So it is a permutation of 4 out of 8 places.
        factorial(6)*(factorial(7)/factorial(7-4))
        ```

+ Question 3
    In how many ways can you arrange the letters of "CAMARADERIE"?

    ```{r}
    # 11 characters.
    # 6 vowels, 5 consonants
    # 3 As, 2 Es, 2 Rs
    ```

    a) Any order.

        ```{r}
        #By the formula of permutation with repetitive letters
        #Assign the value to all_perms object
        all_perms<-factorial(11)/(factorial(3)*factorial(2)*factorial(2))
        all_perms
        ```

    b) Vowels together?

        ```{r}
        #Assume all vowels are single "character" again. So 6 characters
        (factorial(5+1)/(factorial(2)))*(factorial(6)/(factorial(3)*factorial(2)))
        ```

    c) Vowels in alphabetical order?

        ```{r}
        #Same as the last question. But be careful about identical vowels.
        all_perms/(factorial(6)/(factorial(3)*factorial(2)))
        ```


    d) There should be no consecutive vowels?

        ```{r}
        #Same as the last question. But be careful about identical vowels.
        (factorial(5)/factorial(2))*(factorial(6)/factorial(6-6))/(factorial(3)*factorial(2))
        ```

+ Question 4

    Suppose you are putting the top 12 basketball teams in 4 groups evenly (each group should consist of 3 teams). In how many different ways can you arrange the teams?

    ```{r}
    #It is either a chain of combinations or just grouping combination
    choose(12,3)*choose(9,3)*choose(6,3)
    ```

+ Question 5 (20 pts - all equal)

    There are 18 people; 10 from Izmir, 8 from Mugla.

    a) Suppose you want to form a group of 5 people with at least 1 person from Izmir and Mugla. In how many ways can you form such a group?

        ```{r}
        #Calculate as if no rules. It is the combination of 18 to 5.
        #Then remove the combinations of all Izmir or all Mugla people
        choose(18,5) - choose(10,5) - choose(8,5)
        ```

    b) In how many ways can you form a group of 3 people from Izmir and 4 people from Mugla?

        ```{r}
        #Simply separate combinations with multiplication rule.
        choose(10,3)*choose(8,4)
        ```

## Coins, Dice and Cards

When questions mention about coins, dice and cards they are commonly referred items. Nevertheless, you can refer to .

+ Coin tosses: Two possible outcomes. Heads or Tails.
+ Dice rolling: Six possible outcomes. 1-2-3-4-5-6.
+ Card drawing: 52 possible outcomes. There are 4 types (clubs, diamonds, spades and hearts) and 13 ranks for each type. (A)ce - 2 - 3 - 4 - 5 - 6 - 7 - 8 - 9 - 10 - (J)ack - (Q)ueen - (K)ing.



<!--
You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015]. -->
